---
title: "First steps into reproducible bioinformatics"
subtitle: "Seminar at MBIM, Nov 30, 2023"
author:
  - name: Stephan König
    orcid: 0000-0002-1597-3628
    email: stephan.koenig@ubc.ca
    affiliations:
      - name: Skylight, University of British Columbia
        department: Computer Science
        ringgold: 8166
        isni: 0000000122889830
        ror: https://ror.org/03rmrcq20
        
format:
  revealjs:
    theme: [default, ../styles/theme.scss]
    chalkboard: true
    title-slide-attributes:
      data-text-align: right
    
filters:
  - include-code-files
  - ../filters/new-page-links.lua
---

```{r}
#| label: setup

library(conflicted)
library(countdown)
library(dplyr)
conflicts_prefer(dplyr::filter)
library(gt)
library(here)
library(readr)

asv_counts <- read_csv(here("worksheets", "asv-counts.csv"))
```

# Mini lesson:<br>Reproducible reporting of computational analyses {background-color="#faa928"}

## Learning outcomes

-   Describe why combining natural language and code enables reproducible analysis and research.

-   Discover how symbols and text convey formatting and the structure of a document.

## Microbiomes at different lake depths

::: center-xy
Sequence barcode marker of each sample

{{< fa arrow-down title="Down arrow" >}}

Group similar sequences<br>defining<br>amplicon sequence variants (ASVs)<br>ASV ≅ species

{{< fa arrow-down title="Down arrow" >}}

ASV counts table
:::

## Species richness at 10 m depth

::: {layout-ncol="3"}
```{r}
#| label: tbl-asv-counts
#| tbl-cap: ASV counts

asv_counts |>
  gt()
```
:::

```{r}
#| label: asv-counts
#| echo: true
#| output-location: column-fragment

asv_counts |> 
  select(depth_10_m) |>
  filter(depth_10_m >= 1) |> 
  nrow()
```

## How to report the results? {background-image="../images/word.png" background-size="75%" aria-label="An open Word document reads 'The species richness of the sample is 4.'"}

## {{< fa comments >}} Your Turn: Think-pair-share {background-color="#447099"}

How would ensure the correct reporting of your analysis?

`r countdown(minutes = 2)`

::: notes
I would do Think-pair-share here, but due to time constraints, let's get some shoutouts.
:::

## Do you want your grades to be reproducible? {.center .r-fit-text}

## {{< fa file-alt >}} Your Turn: Worksheet {background-color="#447099"}

-   Review the report of a fellow student.
-   Discover how symbols and text structure the document.

::: columns
::: {.column width="50%"}
### Plain text

``` {include="../worksheets/analysis.qmd" filename="analysis.qmd" start-line="1" end-line="7"}
```
:::

::: {.column width="50%"}
### PDF

<iframe src="../worksheets/analysis.pdf" width="100%" height="300">

</iframe>
:::
:::

::: notes
Your worksheet has three pages.
The first one contains the exercises.
The second one has a plain text file printed on a grey background.
The third contains the PDF output.
I would like you to highlight any difference in text or symbols between the two files.
:::

## {{< fa file-alt >}} Your Turn: Worksheet {background-color="#447099"}

-   Take about 2 minute to highlight differences between plain text and the PDF output.
-   Compare your results with your neighbour and work with them on the other questions.

`r countdown(minutes = 5)`

## How Quarto works {background-image="https://cdn.myportfolio.com/45214904-6a61-4e23-98d6-b140f8654a40/c2ef45f1-4a68-4c7a-a43f-bdcf0f90e104_rw_3840.png?h=783eba914284e5adfb9feadcf4ba5f1e" background-size="contain" aria-label="A schematic representing the multi-language input (e.g. Python, R, Observable, Julia) and multi-format output (e.g. PDF, html, Word documents, and more) versatility of Quarto."}

::: footer
Artwork from "Hello, Quarto" keynote by Julia Lowndes and Mine Çetinkaya-Rundel, presented at RStudio Conference 2022.
Illustrated by Allison Horst.
:::

::: notes
Pandoc is a universal translator.
:::

## Markdown

-   Simple expressive formatting with symbols and text.
-   Used in many [apps and services](https://www.markdownguide.org/tools/), for example {{< fa brands discord title="Discord" >}} {{< fa brands slack title="Slack" >}} {{< fa brands github title= "GitHub" >}}.

+-------------------------------------+
| Markdown syntax                     |
+=====================================+
| # \# Heading 1                      |
+-------------------------------------+
| ## \## Heading 2                    |
+-------------------------------------+
| **\*\*strong\*\***                  |
+-------------------------------------+
| *\*emphasized\**                    |
+-------------------------------------+

## {{< fa laptop-code >}} Next lab: Create your own website! {background-color="#447099"}

::: notes
Demo MICB: 2XX website as an example of a Quarto project.
:::

# Universal Design for Learning (UDL):<br>Embracing open source {background-color="#faa928"}

::: notes
[Universal design for learning fellows program](https://tlef.ubc.ca/application/special-call/universal-design-for-learning-fellows-program/)---2024 cohort UDL strives to provide multiple means for three principles of effective instruction.
:::

## Representation

-   Open source tools support alt text for images.
-   Easy to provide headings and descriptive hyperlinks.

## Engagement

-   Follow the **mastery learning** model, i.e., frequent practice, assessment and feedback.
-   Use online assessment platform (e.g., [PrairieLearn](https://prairielearn.readthedocs.io/)) to support mastery learning at scale.
-   Use authentic examples and assessments (for example, ePortfolio)
-   Use relevant, widespread technologies, such as web technologies (HTML, CSS).
-   Model inclusive practices in the way course is run (syllabus) and materials are provided.

## {#prairielearn data-menu-title="PrairieLearn" background-iframe="https://www.prairielearn.com/" background-interactive="true"}

## Action and expression

-   Alternative ways of participating, e.g., discussion section on course website.
-   Increased expression and communication (Quarto offers a lot of flexibility).

##  {#comments data-meny-title="Comments" background-iframe="https://stephan-koenig.github.io/teaching-demo" background-interactive="true"}

# Teaching philosphy: Practicing expert behaviour {background-color="#faa928"}
